## History

```{r, echo = FALSE}
ctbs <- nrow(read.csv(file.path("_data", "epiverse_contributors.csv")))
ctbs_rounded <- floor(ctbs / 10) * 10
# Get the list of repositories in the organization
repos <- gh::gh("GET /orgs/{org}/repos", org = "epiverse-trace")

# Initialize a counter for the number of releases
num_releases <- 0

# Loop over the repositories
for (repo in repos) {
  # Get the list of releases for the repository
  releases <- gh::gh("GET /repos/{owner}/{repo}/releases", owner = repo$owner$login, repo = repo$name)

  # Add the number of releases to the counter
  num_releases <- num_releases + length(releases)
}
```

Founded in 2021, Epiverse-TRACE enables distributed data analysis to power pandemic response. From the start, all the tools we develop are open, collaborative, and intended for real-world impact. Over `r ctbs_rounded` people contributed to `r num_releases` software releases since.

Epiverse-TRACE is part of data.org's Epiverse initiative. In 2021, Wellcome and Rockefeller Foundation enabled London School of Hygiene & Tropical Medicine (LSHTM) and the Medical Research Council Unit The Gambia (MRCG) to join the initiative. In 2022, [Canada’s International Development Research Centre (IDRC)](https://idrc-crdi.ca/en) joined the effort and enabled new collaborators from Latin-America and the Carribean (LAC): Universidad Javeriana and Universidad de los Andes.

Epiverse-TRACE software is used in all kinds of projects. For example, in [projecting epidemics in Gaza](https://web.archive.org/web/20240612123957/https://gaza-projections.org/) (as covered by the [New York Times](https://www.nytimes.com/2024/02/21/health/gaza-war-study-deaths.html)).

## Contributors

We appreciate all contributions throughout Epiverse - big or small 💜 Below you can find anybody who has participated in our discussions, contributed code (in)directly, or reported a bug at some point (in alphabetical order).

Collectively, they make Epiverse possible.

::: {.contributor-grid}

```{r, results='asis', echo = FALSE}
file.path("_data", "epiverse_contributors.csv") |>
  read.csv() |>
  as.list() |>
  purrr::transpose() |>
  purrr::map_chr(~ knitr::knit_expand(file = "_templates/people_element.Rmd",
                                      handle = .x$logins,
                                      avatar = .x$avatar)) |>
  cat(sep = "\n")

```

:::

:::{.callout-caution collapse="true"}

## Did we miss your contribution?

This list is generated automatically every Monday - if you started contributing you should be included soon.

If it's been a while and you're still not showing up - we're sorry 😔 Please let us know on <epiverse@data.org> - we did not mean to exclude you.
:::
