---
title: "Benchmarking as a tool for R package design decision-making"
author:
  - name: "James Mba Azam"
    orcid: "https://orcid.org/0000-0001-5782-7330"
date: "2023/04/25"
categories: [benchmarking, data-driven design, R packages]
format:
  html:
    toc: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

## Motivation

User-facing design decisions in R package development are often with good intent but could have unintended negative impacts on user experience.

In simulation studies, functions are often run multiple times and run times are important. In more advanced set ups, simulation functions are configured to run over multiple cores of a PC or on a High Performance Computer, so differences in speed are not noticeable.

Here, we demonstrate how a simple decision about whether a function should throw a warning or an error in non-fatal events could have a major impact on the function's run time on a computer.

### Example: warnings vs errors

We consider an example where a developer has to decide whether to throw warnings or errors in some non-fatal events.

An example can be found in Epiverse's `bpmodels` package where a function was previously throwing an error for a disallowed combination of arguments but was later [changed to throw a warning](https://github.com/epiverse-trace/bpmodels/pull/53). As part of the change the wrongly specified argument would be fixed internally with a warning to inform the user of what has happened.

This decision was intended to improve user experience. However, it also meant that in a situation where this happens at scale, the warning would be thrown as many times as the number of simulations unless it is fixed.

Throwing warnings can be useful so as to allow the function to run smoothly, especially if an action (for example, certain combinations of arguments are not allowed but not fatal). In a situation where a function throws a warning and is often run multiple times, it would be important that this event does not slow down the the process at scale.

Here, we will demonstrate that such a decision can be driven by data generated from benchmarking.

## Experiment: throwing warnings with base::warning() vs cli::cli_warn()

Let us benchmark a function that uses `cli::cli_warn()` vs `base::warning()`. Here, we choose `cli::cli_warn()` because it has well adopted in the developer community as an alternative to `base::warning()`. Moreover, `cli` helps to provide useful feedback to the user and can be easy and fun to use as a developer when they get a hang of it.

### Setup 1: one arg validation and one warning

We will define a function that takes an input, which is expected to be an integer, else a warning is thrown. For the purpose of this exercise, we assume that a non-integer input is not preferred but is not fatal.

Based on this description, we will need 2 main packages, [cli](https://cli.r-lib.org/articles/semantic-cli.html) for the function in question here and microbenchmark\](https://cran.r-project.org/web/packages/microbenchmark/index.html) to measure the run times. We will also use an optional one, [checkmate](https://cran.r-project.org/web/packages/checkmate/index.html) to validate the input, however, the reader may write their own checks and avoid taking on `checkmate` as a dependency.

Let us load the packages

```{r}
library(microbenchmark)
library(cli)
library(checkmate)
```

#### One `base::warning()` instance

Let us define a function to throw a warning using base R, if an input is not an integer.

```{r}
base_warn_bmrk <- function(x){
  if (!checkmate::testInteger(x)) {
    warning("x must be an integer")
  }
}
```

Let's test `base_warn_bmrk()`.

```{r}
base_warn_bmrk(x = 1.2)
```

As we can see, it works as expected.

Now, let's define a function to throw a warning using {cli_warn}, if an input is not an integer.

#### One `cli::cli_warn()` instance

```{r}
cli_warn_bmrk <- function(x){
  if (!checkmate::testInteger(x)) {
    cli_warn(c(
      "x must be an integer",
      "i" = "x is currently an {.cls {class(x)}}",
      "x" = "x is not an integer"
    )
    )
  }
}
```

Let's test `cli_warn_bmrk()`:

```{r}
cli_warn_bmrk(x = 1.2)
```

`cli_warn_bmrk()` also works as expected.

We will also check if the benchmark scales with the number of warnings in the function. To do this, we will modify the original functions to take two inputs, both of which are expected to be integers and follow the same logic as before.

### Setup 2: two argument validation and two warnings

#### Two `base::warning()` instances

Let us define a function to throw a warning using base R, if an input is not an integer.

```{r}
base_warn_bmrk2 <- function(x, y){
  # Validate arg x
  if (!checkmate::testInteger(x)) {
    warning("x must be an integer")
  }
  
  # Validate arg y
  if (!checkmate::testInteger(y)) {
    warning("y must be an integer")
  }
}
```

Let's test `base_warn_bmrk2()`.

```{r}
base_warn_bmrk2(x = 1.2, y = 1.23)
```

`base_warn_bmrk2` works as expected.

#### Two `cli::cli_warn()` instances

```{r}
cli_warn_bmrk2 <- function(x, y){
  # Validate arg x
  if (!checkmate::testInteger(x)) {
    cli_warn(c(
      "x must be an integer",
      "i" = "x is currently an {.cls {class(x)}}",
      "x" = "x is not an integer"
    )
    )
  }
  
  # Validate arg y
    if (!checkmate::testInteger(y)) {
    cli_warn(c(
      "y must be an integer",
      "i" = "y is currently an {.cls {class(x)}}",
      "x" = "y is not an integer"
    )
    )
  }
  
}
```

Let's test `cli_warn_bmrk2()`:

```{r}
cli_warn_bmrk2(x = 1.2, y = 1.25)
```

`cli_warn_bmrk2()` also works as expected.

Now, to the fun part, let us run the functions $1000$ times and plot the results.

## Results

### The case of 1 arg validation and 1 warning

Here, as we know the functions will throw warnings, we will suppress them to avoid cluttering this article with their outputs.

```{r warning=FALSE}

reps <- 1000

input_vec <- 1.2 

bnmrk_df <- microbenchmark(
  cli_warn = cli_warn_bmrk(x = input_vec),
  base_warn = base_warn_bmrk(x = input_vec),
  times = reps
)

boxplot(bnmrk_df)
title(paste("Function with 1 argument of length 1 and throws 1 warning (replicated", reps, 'times)'))
```

Let us try this with a mixed vector of length $10$ and see how that scales up.

```{r warning=FALSE}
set.seed(123)

reps <- 1000

input_vec2 <- c(runif(5, 0, 10), sample.int(6, 5))

bnmrk_df2 <- microbenchmark(
  cli_warn = cli_warn_bmrk(x = input_vec2),
  base_warn = base_warn_bmrk(x = input_vec2),
  times = reps
)

boxplot(bnmrk_df2)
title(paste("Function with 1 argument of length 10 and throws 1 warning (replicated", reps, 'times)'))
```

### The case of 2 arg validation and 2 warnings

```{r warning=FALSE}

reps <- 1000

arg_x <- 1.2 

arg_y <- 1.25

bnmrk_df3 <- microbenchmark(
  cli_warn = cli_warn_bmrk2(x = arg_x, y = arg_y),
  base_warn = base_warn_bmrk2(x = arg_x, y = arg_y),
  times = reps
)

boxplot(bnmrk_df3)

title(paste("Function with 2 arguments of length 1 each and throws 2 warning (replicated", reps, 'times)'))
```

We can see from this simple example that `cli::cli_warn()` was $100$ times slower on average than `base::warn()`. Additionally, the benchmark was not affected by the length of the input but rather scaled linearly with the number of calls to the warning functions.

## Conclusion

In designing R package infrastructure, it is important that the developer(s) considers the impact of user-facing decisions. Here, we have demonstrated how this decision can be driven with data from benchmarking exercises. We show how a simple decision to use `cli::cli_warn()` to throw warnings could affect function run times.

The demonstration here can be extended to other design decisions. We recommend benchmarking as a way to help the developer make design decisions.
