---
title: "Using benchmarking to guide the adoption of dependencies in R packages"
author:
  - name: "James Mba Azam"
    orcid: "https://orcid.org/0000-0001-5782-7330"
date: last-modified
categories: [benchmarking, R, cli, R packages, software architecture]
bibliography: index.bib
format:
  html:
    toc: true
execute: 
  cache: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  message = FALSE
)
```

R package developers often have to take on dependencies for various reasons.
Dependencies are external packages that provide extra functionalities or
features to another package [@wickham2023r]. Quite often, the goal of taking
on a dependency is to utilize its rich features and reduce duplicated coding
effort. For example, the `ggplot2` package is often adopted, instead of R's
inbuilt plotting functions, to make visualizations using its declarative
system and implementation of the Grammar of Graphics, and slick automation
across the board.

Sometimes, introducing a dependency might slow down existing processes. Hence,
it is important to consider the speed of a dependency's functions before
adopting it. Code optimisation is often reserved for heavy processes like data
processing steps. However, here, we will focus on lesser optimised processes,
and in particular, printing of texts in R.

For this demonstration, we will look at a situation where a package developer
is considering options for condition signalling beyond the functionalities
provided in base R. Condition signalling refers to the process of communicating
to a user that an issue has arisen during code execution [@wickham2019advanced].

Base R ships with functionalities for condition signalling. There is
`base::stop()` for signalling errors.  Errors refer to cases where something fatal
happens during code execution so that it halts, for example, when a function is
run without specifying the required arguments.

```{r}
try(rnorm())
```

Base R provides `base::warning()` for throwing warnings. Warnings are used to
handle less fatal issues with code execution that do not
need to halt the process. For example, when you compare two vectors of
different lengths, R will throw a warning but will still return a result.

```{r}
c(1:2) > c(1:3)
```

Lastly, there is `base::message()` for throwing messages. Messages are used
to provide useful information about processes. For example, sometimes, when
you load a package, R will throw a message to indicate that the package has
been loaded.

```{r}
#| message: true
library(dplyr)
```

These base R functions are, however, not flexible enough for advanced
formatting: colour, progress bar, contextual pluralisation, and so forth. 
The package developer might want to adopt external dependencies for these
flexibilities. The [cli](https://cli.r-lib.org/articles/semantic-cli.html)
package is well designed for this. `cli` offers many of the advanced formatting
features. Here is a quick example (adopted from the help file) using
`cli::cli_abort()`, the equivalent of `base::stop()`.

```{r}
n <- "c(1, 2, 3)"
try(
  cli_abort(
    c(
      "{.var n} must be a numeric vector",
      "x" = "You've supplied a {.cls {class(n)}} vector."
    )
  )
)
```

In this example, we used inline text formatting to print `n` as a variable
and interpolated code to print the class of `n` in the error message and all in
just 6 lines of code.   

If this is enough to excite your interests, check out `cli`'s [website](https://cli.r-lib.org/)
and help files, which provide in-depth guides on its myriad features.

Now, considering all these amazing features that `cli` offers, what should a
package developer consider to adopt it? In this article, we will
demonstrate how to use (micro)benchmarking[^1] to guide the decision through
the lense of speed. Benchmarking can be used to check whether
adopting `cli` as a dependency would slow down existing printing
processes in undesirable ways.

Here, we will benchmark `cli::cli_warn()` against base R's `base::warning()` to
see if the former has any speed issues. The results of this exercise will
provide us with insights on whether its worth taking on `cli` for its features
as a trade-off for slower printing speed.

## Experiment

We will consider different scenarios of number of warnings thrown to
tease out the relationship between speed of the function and number of warnings
thrown in a function with and without `cli::cli_warn()`. We will also check how
the benchmark scales with the number of warnings in the function.

Based on this description, we will need the following packages: `cli` for its
warning function, and [bench](https://cran.r-project.org/web/packages/bench/index.html)
for measuring the run times,  [dplyr](https://dplyr.tidyverse.org/)
for manipulating the results, and [ggplot2](https://ggplot2.tidyverse.org/)
for plotting.

```{r}
library(bench)
library(cli)
library(dplyr)
library(ggplot2)
```

Let's define a function that takes an argument `n` for the number of warnings
to throw and `pkg` for the type of package to use.

```{r}
warn_Ntimes <- function(n, pkg) {
  warning_msg <- "x must be an integer"
  switch(pkg,
    base = for (i in seq_len(n)) {
      warning(warning_msg)
    },
    cli = for (i in seq_len(n)) {
      cli_warn(warning_msg)
    }
  )
}
```

Let's test our function to see if it works as expected.

```{r}
warn_Ntimes(3, "base")
warn_Ntimes(3, "cli")
```

Now, we'll consider scenarios where a function throws 1, 5, 10, 15, 20, and
100 warnings using base R and `cli`.

```{r warning=FALSE}
# Number of warnings to throw
warnings <- c(1, 5, 10, 15, 20, 100)
```

Let's run benchmarks over the different scenarios and store results
in a data.frame.

```{r warning=FALSE,message=FALSE}
bnmark_res <- bench::press(
  warnings = c(1, 5, 10, 15, 20, 100),
  bench::mark(
    cli = warn_Ntimes(warnings, "cli"),
    base = warn_Ntimes(warnings, "base")
    )
  )
```

## Results

It's time to explore the results of the data generated. Let's first make a
boxplot of the run times for the different scenarios.

Drum roll please...

```{r}
autoplot(
  bnmark_res,
  type = "boxplot"
  ) +
  labs(
    x = "Package",
    caption = "Functions throwing various numbers of warnings as indicated in the facet label"
  )
```

Now, let's see how the benchmarks scale with the number of warnings thrown.

```{r}
bnmark_res$pkg <- attr(bnmark_res$expression, "description")

ggplot(bnmark_res) +
  geom_line(
    aes(
      x = warnings,
      y = median,
      color = pkg
    )
  )
```

As we can see, `cli` is consistently slower than base R for the same number of
warnings thrown. The median times also follow a similar non-linear trajectory.
Benchmarking experiments with other `cli` functions have revealed similar non-linear
relationships between its speed and the number of times it invokes those
functions [^2]. However, those exercises were not compared with equivalent functions
in base R. The relative comparison here is useful for decision-making.

The developers of `cli` have also conducted benchmarks of the `ansi_*()` family
of functions in `cli` in comparison to base R and the `fansi` package.
They find that `cli` is consistently slower than base R, which corroborates the
results of our experiment here. These benchmarks are available in the `cli`
documentation [^3].

So, should we be worried about the speed of `cli`? Well, it depends on the context.
The "R Packages" book by Hadley Wickham and Jenny Bryan suggests approaching
such a decision from a holistic, balanced, and quantitative
approach[@wickham2023r]. We'll leave the reader to make their own decision
based on their use case.

Most developers might argue that this is an optimisation overkill[^4]. However,
it is important to consider speed differences in context.
In the case of simple printing, the speed difference is
negligible yet disruptive and somewhat painful. However, in the grand scheme
of things, this might be nothing compared with much slower processes that
need more attention. In those cases, the developer might want to consider other
optimisation strategies such as profiling [^5]. The essence of this experiment
is to demonstrate the utility of benchmarking in making quick design decisions.

## Conclusion

In designing R package infrastructure with dependencies, it might
sometimes be necessary to check if they don't slow down existing processes.
Here, we have demonstrated how benchmarking is one way to achieve
this for a process involving condition signalling. We show how a simple
decision to use `cli::cli_warn()` to handle warnings could come at the cost of
a tiny loss in speed, which is worth considering in its context.

The demonstration here can be extended to other dependency adoption decisions for
input checking, loops, object manipulations, and so forth. We recommend benchmarking
as a way to help developers make quick design decisions. However, we also recommend
that developers consider the context of the optimisation in interpreting the results.

## Other R packages for benchmarking

* [microbenchmark](https://github.com/joshuaulrich/microbenchmark): an R
  package for comparing the execution time of R expressions.
* [rbenchmark](https://code.google.com/archive/p/rbenchmark/): an R package
  for benchmarking R code.
* [tictok](https://github.com/collectivemedia/tictoc): an R package to time
R functions
* [touchstone](https://github.com/lorenzwalthert/touchstone/tree/main): an R
  package for
  benchmarking of pull requests with statistical confidence.

[^1]: Benchmark (Wikipedia): <https://en.wikipedia.org/wiki/Benchmark_(computing)>
[^2]: Formating errors can be slow/variable: <https://github.com/r-lib/cli/issues/617>
[^3]: cli benchmarks: <https://cli.r-lib.org/articles/ansi-benchmark.html>
[^4]: Donald Knuth's quoted as having said, "The real problem is that programmers have spent far too much time worrying about efficiency in the wrong places and at the wrong times; premature optimization is the root of all evil (or at least most of it) in programming." (See <https://en.wikiquote.org/wiki/Donald_Knuth>)
[^5]: Profiling (Wikipedia): <https://csgillespie.github.io/efficientR/performance.html>
